Computing solution space properties of combinatorial optimization problems by generic programming tensor networks \thanks{\funding{...}}

### 2022-04-12 paper/paper.tex
Updates from Overleaf

\begin{abstract}
%We introduce a method that uses generic programming tensor networks to compute various properties of a class of hard graph problems that can be decribed by local constraints.
%We introduce a tensor network algorithm %a method that combines generic programming and tensor network techniques 
We introduce a unified framework to compute the solution space properties of a class of combinatorial optimization problems including the independent set problem, the maximum cut problem, the vertex coloring problem, the clique problem, the dominating set problem, the set packing problem, the set covering problem, and the satisfiability problem.
%Let the independent set problem be an example, we show how to compute, for example, the maximum several independent set sizes, the counting of independent sets of a give size, the enumeration of independent sets of a given size.
These solution space properties include, for example, finding one of the optimum solutions, counting the number of solutions of a given size, and enumeration and sampling of solutions of a given size.
We use the independent set problem as an example and show how all these solution space properties can be computed in the unified approach of generic programming tensor networks.
%By making use of generic programming, our algorithms are very simple to implement and one can in addition directly utilize recent advances in tensor network contraction techniques such as near-optimal contraction order finding and slicing to achieve high performance.
%By using generic programming techniques, the same framework can be used to compute all of these properties. 
%Our algorithm utilizes recent advances in tensor network contraction techniques to achieve high performance, including methods to quickly find a near-optimal contraction order and slicing.
%The algorithmic complexity of this approach is two to the treewidth of the line graph of the tensor network graph.
To demonstrate the versatility of this computational tool, we apply it to a few examples, including computing the entropy constant for hardcore lattice gases, studying the overlap gap properties, and analyzing the performance of quantum and classical algorithms for finding maximum independent sets.
% including the calculations of the entropy constant for some hardcore lattice gases on 2D square lattices and the study of the overlap gap property on three regular graphs.
%We show how our method helps to understand these hard problems on systems with hundreds of vertices.
\end{abstract}

@@

In this paper, we introduce a unified framework to compute a broader class of quantities associated with these problems, beyond just finding an optimum solution.
We call them \textit{solution space properties}, which can be much harder to compute (e.g.\ \#P-complete), but they can be crucial for understanding hard combinatorial optimization problems.
% In this paper, we focus on the less studied aspects of computing \textit{solution space properties}, which can be much harder to compute, %; for example, counting independent sets of a general graph is \#P-complete~\cite{Vadhan2001}, 
% but they can be crucial towards understanding hard combinatorial optimization problems. 
These \textit{solution space properties} can include not only the maximum or minimum set size but also the number of sets at a given size, enumeration of all sets at a given size, and direct sampling of such sets when they are too large to be fit into memory.
% For the weighted version of the above problems, it also includes finding the largest or smallest $k$ sets and their sizes.
%are many interesting and hard computational problems concerning various properties of independent sets.
%Solution space property computation allows us to understand the solution space better.
They can be used to understand the hardness of finding an optimum solution for a problem instance and the performance of a solver in computing the solution for the given instance.
For example, the number of configurations at different sizes can inform how likely a simulated annealing algorithm will be trapped in local minima at certain sizes~\cite{Xu2018}.
%In the past, people use the counting of configurations at different sizes to tell how likely the simulated annealing will be trapped in certain size in the strong defensive alliance problem~\cite{Xu2018}, 
The pair-wise Hamming distance distribution of configurations at a given size can indicate the presence or absence of the overlap gap property~\cite{Gamarnik2013, Gamarnik2019}, which can be used to bound the performance of local search algorithms.
%a no-go theorem about when the solution space geometry is formidably
%hard for any local search algorithm.
In a recent experiment based on a Rydberg atom array quantum computer, the authors use the counting and the configuration space connectivity information to find maximum independent set (MIS) problem instances that are hard for simulated annealing \cite{Ebadi2022}.
% In a recent experiment using a Rydberg atom array quantum simulator to find maximum independent sets (MIS) of a graph~\cite{Ebadi2022}, the authors use the counting information and the configuration space connectivity to find instances hard for simulated annealing. %and design better quantum algorithms 
%to solve these instances faster, 
The need to understand this experiment motivates us to find methodologies to compute these solution space properties.
%\purple{I would give examples of why we want to compute solution space properties.
%For instance, in statistical physics, computing thermal states and
%observables via the partition function. In complexity theory, proving limits on local algorithms (OGP).}\blue{now?}

@@

Let us characterize the independent set problem on graph $G=(V, E)$ as an energy model with two parts
\begin{equation}\label{eq:eng}
    \mathcal{E}(G, s) = -\sum_{i\in V} w_i s_i + \infty \sum_{\langle i,j\rangle \in E}s_i s_j
\end{equation}
where $s_i$ is a spin on vertex $i \in V$ and $w_i$ is an onsite energy term associated with it.
The first part corresponds to the negative independent set size and the second part describes the independence constraint, which corresponds to the Rydberg blockade~\cite{Pichler2018, Ebadi2022} in cold atom arrays or the repulsive force in hardcore lattice model~\cite{Dyre2016, Fernandes2007}.
The partition function is defined as
\begin{equation}\label{eq:partition}
    \begin{split}
    Z(G, \beta) = \sum_{s}e^{-\beta \mathcal{E}(G, s)} = \sum_{s\in \mathcal{I}(G)} e^{\beta \sum w_i s_i}\\
         = \sum_{k=0}^{\alpha(G)}a(k) e^{\beta k}  \qquad \quad (k = \sum w_i s_i)
    \end{split}
\end{equation}
where $\mathcal{I}(G)$ is the set of independent sets of graph $G$, $\alpha(G)$ is the absolute value of the minimum energy (maximum independent set size), $a(k)$ is the number of spin configurations with energy $-k$ (independent sets of size $k$).
The partition function can be expressed as a tensor network by placing a vertex tensor on each spin $i$
\begin{equation}
    W(e^{\beta w_i}) = \left(\begin{matrix}
        1 \\
        e^{\beta w_i}
    \end{matrix}\right),
\end{equation}
and an edge tensor on each bond
\begin{equation}
       B = \left(\begin{matrix}
        1  & 1\\
        1 & 0
    \end{matrix}\right),
\end{equation}
where the $0$ in the edge tensor comes from $e^{-\beta\infty}$ in the second term of \Eq{eq:eng}.
By letting $x = e^{\beta}$, we get the tensor network for computing the independence polynomial as described by \Eq{eq:vertextensor} and \Eq{eq:edgetensor}.
If we further let $w_i=1$, the second line of \Eq{eq:partition} is equivalent to the independence polynomial.

\section{An example of contraction complexity increase due to using the standard tensor network notation}\label{sec:tensorbad}

@@

In the following example, we will show this conversion might increase the contraction complexity of a tensor network.
Let us consider the following King's graph.

@@

\centerline{\begin{tikzpicture}
    \def\r{0.08}
    \def\a{0.07}
    \def\L{0.6}
    \def\l{0.1}
    \def\sql{0.24}
    \pgfmathsetseed{2}
    \foreach[evaluate={\cr=0.1+0.5*Mod(\x,2)}] \x in {1,...,5}
        \foreach[evaluate={\cg=0.1+0.3*Mod(\y,2); \cy=0.5-0.5*Mod(\y,2)}] \y in {1,...,5}{
            \edef\R{\pdfuniformdeviate 255}
            \edef\G{\pdfuniformdeviate 255}
            \edef\B{\pdfuniformdeviate 255}
            \xdefinecolor{MyColor}{RGB}{\R,\G,\B}
            \ifnum \x < 5
                \draw [thick, MyColor, opacity=1.0, line cap=round] (\x,\y) -- (\x+0.5,\y);
                \ifnum \y < 5
                \draw [thick, MyColor, opacity=1.0, line cap=round] (\x,\y) -- (\x+0.3,\y+0.3);
                \fi
                \ifnum \y > 1
                \draw [thick, MyColor, opacity=1.0, line cap=round] (\x,\y) -- (\x+0.3,\y-0.3);
                \fi
            \fi
            \ifnum \x > 1
                \draw [thick, MyColor, opacity=1.0, line cap=round] (\x,\y) -- (\x-0.5,\y);
                \ifnum \y < 5
                \draw [thick, MyColor, opacity=1.0, line cap=round] (\x,\y) -- (\x-0.7,\y+0.7);
                \fi
                \ifnum \y > 1
                \draw [thick, MyColor, opacity=1.0, line cap=round] (\x,\y) -- (\x-0.7,\y-0.7);
                \fi
            \fi
            \ifnum \y < 5
                \draw [thick, MyColor, opacity=1.0, line cap=round] (\x,\y) -- (\x,\y+0.5);
            \fi
            \ifnum \y > 1
                \draw [thick, MyColor, opacity=1.0, line cap=round] (\x,\y) -- (\x,\y-0.5);
            \fi
        }
    \foreach \x in {1,...,5}
        \foreach \y in {1,...,5}{
            %\filldraw[fill=black] (\x,\y) circle [radius=0.7*\r];
        }
    \foreach \x in {1,...,5}
        \foreach \y in {1,...,5}{
            \ifnum \y < 5
                \filldraw[fill=black] (\x,\y+0.5) circle [radius=\r];
                \filldraw[fill=black] (\y+0.5,\x) circle [radius=\r];
            \fi
        }
    \foreach \x in {1,...,4}
        \foreach \y in {1,...,4}{
            \filldraw[fill=black] (\x+0.3,\y+0.3) circle [radius=\r];
            \filldraw[fill=black] (\y+0.3,\x+0.7) circle [radius=\r];
        }
    \tikzset{decoration={snake,amplitude=.4mm,segment length=2mm,
                    post length=0mm,pre length=0mm}}
    \draw [decorate] (2.7, 0.5) -- (2.7, 5.5);
\end{tikzpicture}}
Vertex tensors are not shown here because they can be absorbed into an edge tensor hence do not change the contraction complexity.
If we contract this tensor network in the column-wise order, the maximum intermediate tensor has rank $\sim L$, which can be seen by counting the number of colors at the cut.

@@

In this diagram, the additional $\delta$ tensors can have ranks up to $8$.
If we still contract this tensor network in a column-wise order, the maximum intermediate tensor has rank $\sim3L$, i.e. the space complexity is $\approx 2^{3L}$.

\section{Hard problems and tensor networks}\label{sec:otherproblems}
\subsection{Maximal independent sets and maximal cliques}\label{sec:maximal}
In this section, we focus the discussion on the maximal independent sets problem since finding maximal cliques of a graph is equivalent to finding the maximal independent sets of the complement of the graph.
Let $G=(V,E)$ be a graph, we denote the neighborhood of a vertex $v\in V$ as $N(v)$.
A maximal independent set $I_m$ is an independent set that there does exist a vertex $v \in V$ such that $I_m \cap (\{v\} \cup N[v])  = \emptyset$, i.e. an independent set that can not become a larger one by adding a new vertex.
To characterize the maximal independence restriction, we defined a tensor on each $\{v\}\cup N(v)$ as
\begin{equation}\label{eq:maximal}
    T(x_v, w_v)_{s_1,s_2,\ldots,s_{|N(v)|},s_v} = \begin{cases}
        s_vx_v^{w_v} & s_1=s_2=\ldots=s_{|N(v)|}=0,\\
        1-s_v& \text{otherwise}.\\
    \end{cases}
\end{equation}
It means if none of the neighbours of $v$ are in $I_{m}$ ($ s_1=s_2=\ldots=s_{|N(v)|}=0$), then $v$ must be in $I_{m}$ and contribute a factor $x_{v}$, otherwise, if any of the neighbourhood vertices is in $I_{m}$, then $v$ cannot be in $I_{m}$.
For a vertex $v$ having degree 2, this tensor has the following form
\begin{equation}
    T(x_v, w_v)=\left(\begin{matrix}
    \left(\begin{matrix}
        ~~0 &~1 \\
        ~~1 &~1
    \end{matrix}\right)\\
    \left(\begin{matrix}
        x_v^{w_v} &0 \\
        0 &0
    \end{matrix}\right)
    \end{matrix}\right).
\end{equation}

Let us consider the graph in Example~\ref{eg:tensorcontraction}, the corresponding tensor network structure for computing the maximal independent polynomial has the following hypergraph representation.

@@

%The contraction complexity of this tensor network is often greater than the one for the independence set problem.
By contracting this tensor network with generic element types,
we can compute the maximal independent set properties such as the maximal independence polynomial and the enumeration of maximal independent sets.
The maximal independence polynomial is defined as
\begin{equation}
D_{m}(G, x) = \sum_{k=0}^{\alpha(G)} b_k x^k,
\end{equation}
where $b_k$ is the number of maximal independent sets of size $k$.
Comparing with the independence polynomial in \Eq{eq:idpdef}, we have $b_{k} \leq a_{k}$ and $b_{\alpha(G)} = a_{\alpha(G)}$. $D_m(G, 1)$ counts the total number of maximal independent sets~\cite{Gaspers2012, Manne2013},
which to our knowledge, the best algorithm gives a time complexity $O(1.3642^{|V|})$~\cite{Gaspers2012}.
%If we want to find an MIS, $b_{k}$ can, in some cases, provide hints on the difficulty of finding the MIS using local algorithms~\cite{Ebadi2022}.
%The uni-modality, log-concavity, and real-rootness properties of the maximal independence polynomial for special classes of graphs have also been studied~\cite{Hu2017}.

We show the benchmark of computing the maximal independent set properties on $3$-regular graphs in \Fig{fig:benchmark-maximal},
including a comparison to the Bron-Kerbosch algorithm from Julia package \href{https://github.com/JuliaGraphs/Graphs.jl}{Graphs}~\cite{Graphs}.
\Fig{fig:benchmark-maximal} (a) shows the space and time complexities of tensor contraction, which are typically larger than those for the independent set problem.
In \Fig{fig:benchmark-maximal} (b), one can see counting maximal independent sets are much more efficient than enumerating them, while our generic tensor network approach runs slightly faster than the Bron-Kerbosch approach in enumerating all maximal independent sets.

\begin{figure} 
    \centering
    \includegraphics[width=\textwidth, trim={0cm 0cm 0cm 0cm}, clip]{figures/fig2.pdf}
    \caption{Benchmarks of computing different solution space properties of the maximal independent sets (ISs) problem on random three regular graphs at different sizes.
    (a) time and space complexity of tensor network contraction. 
    (b) The wall clock time for counting and enumeration of maximal ISs. The missing data is due to due to the memory limit.
    }
    \label{fig:benchmark-maximal}
\end{figure}

@@

\subsection{Vertex coloring}
Let $G=(V,E)$ be a graph, a vertex coloring is an assignment of colors to each vertex $v\in V$ such that no edge connects two identically colored vertices. 
In a $k$-coloring problem, the number of colors is limited to less or equal to $k$.
Let us use the 3-coloring problem as an example to show how to reduce it to tensor contractions.
We first map a vertex $v \in V$ to a degree of freedom $c_v\in\{0,1,2\}$.
Then we define a tensor labeled by $c_v$ for each vertex $v$ as
\begin{equation}
    W = \left(\begin{matrix}
        r_v\\
        g_v\\
        b_v
    \end{matrix}\right),
\end{equation}
and a bond tensor labelled by $(c_u, c_v) $ for each edge $(u, v) \in E$ as
\begin{equation}
    B(x^{w_{uv}}) = \left(\begin{matrix}
        0 & x^{w_{uv}} & x^{w_{uv}}\\
        x^{w_{uv}} & 0 & x^{w_{uv}}\\
        x^{w_{uv}} & x^{w_{uv}} & 0
    \end{matrix}\right),
\end{equation}
where $r_v$, $g_v$ and $b_v$ are colors for labeling the configurations. $B$ tensors are for specifying the coloring constraints and $W$ tensors are for labeling the solutions.
%The number of possible colorings can be obtained by contracting this tensor network by setting $r_v, g_v$ and $b_v$ to $1$.
Let $x^{w_{uv}} = x$ and $r_v = g_v = b_v = 1$, we have a graph polynomial,
in which the $k$-th coefficient is the number of coloring that $k$ bonds satisfy constraints.
If a graph is colorable, the maximum order of this polynomial should be equal to the number of edges in this graph.
Similarly, one can define an edge coloring problem by defining the tensor network on the line graph of $G$.

\subsection{Cutting problem}
In graph theory, a cut is a partition of the vertices of a graph into two disjoint subsets,
which is also known as the spin-glass problem in statistic physics.
Let $G=(V,E)$ be a graph, we associate a weight $w_v$ to each $v\in V$. To reduce the cutting problem on $G$ to the contraction of a tensor network, we first define a Boolean degree of freedom $s_v\in\{0, 1\}$ for each vertex $v\in V$.
Then for each edge $(u,v)\in E$, we define an edge matrix labelled by $(s_u, s_v)$ as
\begin{equation}
    B(x^{w_{uv}}_u, x^{w_{uv}}_v) = \left(\begin{matrix}
        1 & x_{v}^{w_{uv}}\\
        x_{u}^{w_{uv}} & 1
    \end{matrix}\right),
\end{equation}
where variables $x_u^{w_{uv}}$ and $x_v^{w_{uv}}$ are for a cut on this edge or a domain wall in a spin-glass.
Let $x_u^{w_{uv}} = x_v^{w_{uv}} = x$, we have a graph polynomial similar to previous ones,
in which the $k$th coefficient is two times the number of cut configurations having size $k$ (i.e. cutting $k$ edges).

>>>>

\begin{abstract}
%We introduce a method that uses generic programming tensor networks to compute various properties of a class of hard graph problems that can be decribed by local constraints.
%We introduce a tensor network algorithm %a method that combines generic programming and tensor network techniques 
We introduce a unified framework to compute the solution space properties of a class of combinatorial optimization problems including the independent set problem, the maximum cut problem, the vertex coloring problem, the clique problem, the dominating set problem, the set packing problem, the set covering problem, and the satisfiability problem.
%Let the independent set problem be an example, we show how to compute, for example, the maximum several independent set sizes, the counting of independent sets of a give size, the enumeration of independent sets of a given size.
These solution space properties include finding one of the optimum solutions, counting the number of solutions of a given size, and enumeration and sampling of solutions of a given size.
We use the independent set problem as an example and show how all these solution space properties can be computed in the unified approach of generic programming tensor networks.
%By making use of generic programming, our algorithms are very simple to implement and one can in addition directly utilize recent advances in tensor network contraction techniques such as near-optimal contraction order finding and slicing to achieve high performance.
%By using generic programming techniques, the same framework can be used to compute all of these properties. 
%Our algorithm utilizes recent advances in tensor network contraction techniques to achieve high performance, including methods to quickly find a near-optimal contraction order and slicing.
%The algorithmic complexity of this approach is two to the treewidth of the line graph of the tensor network graph.
To demonstrate the versatility of this computational tool, we apply it to several  examples, including computing the entropy constant for hardcore lattice gases, studying the overlap gap properties, and analyzing the performance of quantum and classical algorithms for finding maximum independent sets.
% including the calculations of the entropy constant for some hardcore lattice gases on 2D square lattices and the study of the overlap gap property on three regular graphs.
%We show how our method helps to understand these hard problems on systems with hundreds of vertices.
\end{abstract}

@@

In this paper, we introduce a unified framework to compute a broad class of quantities associated with these problems, beyond just finding an optimum solution.
We call them \textit{solution space properties}. In practice these can be much harder to compute (corresponding e.g.\ to \#P-complete class). However, these properties can be crucial for understanding detailed properties of hard combinatorial optimization problems.
% In this paper, we focus on the less studied aspects of computing \textit{solution space properties}, which can be much harder to compute, %; for example, counting independent sets of a general graph is \#P-complete~\cite{Vadhan2001}, 
% but they can be crucial towards understanding hard combinatorial optimization problems. 
These \textit{solution space properties} can include not only the maximum or minimum set size but also the number of sets at a given size, enumeration of all sets at a given size, and direct sampling of such sets when they are too large to be fit into memory.
% For the weighted version of the above problems, it also includes finding the largest or smallest $k$ sets and their sizes.
%are many interesting and hard computational problems concerning various properties of independent sets.
%Solution space property computation allows us to understand the solution space better.
They can be used to understand the hardness of finding an optimum solution for a problem instance and the performance of a solver in computing the solution for the given instance.
For example, the number of configurations at different sizes can inform how likely a simulated annealing algorithm will be trapped in local minima at certain sizes~\cite{Xu2018}.
%In the past, people use the counting of configurations at different sizes to tell how likely the simulated annealing will be trapped in certain size in the strong defensive alliance problem~\cite{Xu2018}, 
The pair-wise Hamming distance distribution of configurations at a given size can indicate the presence or absence of the overlap gap property~\cite{Gamarnik2013, Gamarnik2019}, which can be used to bound the performance of local search algorithms.
%a no-go theorem about when the solution space geometry is formidably
%hard for any local search algorithm.
In a recent experiment based on a Rydberg atom array quantum computer, the authors use the counting and the configuration space connectivity information to find maximum independent set (MIS) problem instances that are hard for simulated annealing \cite{Ebadi2022}.
% In a recent experiment using a Rydberg atom array quantum simulator to find maximum independent sets (MIS) of a graph~\cite{Ebadi2022}, the authors use the counting information and the configuration space connectivity to find instances hard for simulated annealing. %and design better quantum algorithms 
%to solve these instances faster, 
The need to understand this experiment motivates us to find methodologies to compute these solution space properties.
%\purple{I would give examples of why we want to compute solution space properties.
%For instance, in statistical physics, computing thermal states and
%observables via the partition function. In complexity theory, proving limits on local algorithms (OGP).}\blue{now?}

@@

Let us characterize the independent set problem on graph $G=(V, E)$ as an energy model with two parts
\begin{equation}\label{eq:eng}
    \mathcal{E}(G, s) = -\sum_{i\in V} w_i s_i + \infty \sum_{\langle i,j\rangle \in E}s_i s_j
\end{equation}
where $s_i$ is a spin on vertex $i \in V$ and $w_i$ is an onsite energy term associated with it.
The first part corresponds to the negative independent set size and the second part describes the independence constraint, which corresponds to the Rydberg blockade~\cite{Pichler2018, Ebadi2022} in cold atom arrays or the repulsive force in hardcore lattice models~\cite{Dyre2016, Fernandes2007}.
The partition function is defined as
\begin{equation}\label{eq:partition}
    \begin{split}
    Z(G, \beta) = \sum_{s}e^{-\beta \mathcal{E}(G, s)} = \sum_{s\in \mathcal{I}(G)} e^{\beta \sum w_i s_i}\\
         = \sum_{k=0}^{\alpha(G)}a(k) e^{\beta k}  \qquad \quad (k = \sum w_i s_i)
    \end{split}
\end{equation}
where $\mathcal{I}(G)$ is the set of independent sets of graph $G$, $\alpha(G)$ is the absolute value of the minimum energy (maximum independent set size), $a(k)$ is the number of spin configurations with energy $-k$ (independent sets of size $k$).
The partition function can be expressed as a tensor network by placing a vertex tensor on each spin $i$
\begin{equation}
    W(e^{\beta w_i}) = \left(\begin{matrix}
        1 \\
        e^{\beta w_i}
    \end{matrix}\right),
\end{equation}
and an edge tensor on each bond
\begin{equation}
       B = \left(\begin{matrix}
        1  & 1\\
        1 & 0
    \end{matrix}\right),
\end{equation}
where the $0$ in the edge tensor comes from $e^{-\beta\infty}$ in the second term of \Eq{eq:eng}, which is the independence constraint.
By letting $x = e^{\beta}$, we get the tensor network for computing the independence polynomial as described by \Eq{eq:vertextensor} and \Eq{eq:edgetensor}.
If we further let $w_i=1$, the second line of \Eq{eq:partition} is equivalent to the independence polynomial.

\section{An example of increased contraction complexity for the standard tensor network notation}\label{sec:tensorbad}

@@

In the following example, we will show this conversion might increase the contraction complexity of a tensor network.
Let us consider the following King's graph. \\

@@

\centerline{\begin{tikzpicture}
    \def\r{0.08}
    \def\a{0.07}
    \def\L{0.6}
    \def\l{0.1}
    \def\sql{0.24}
    \pgfmathsetseed{2}
    \foreach[evaluate={\cr=0.1+0.5*Mod(\x,2)}] \x in {1,...,5}
        \foreach[evaluate={\cg=0.1+0.3*Mod(\y,2); \cy=0.5-0.5*Mod(\y,2)}] \y in {1,...,5}{
            \edef\R{\pdfuniformdeviate 255}
            \edef\G{\pdfuniformdeviate 255}
            \edef\B{\pdfuniformdeviate 255}
            \xdefinecolor{MyColor}{RGB}{\R,\G,\B}
            \ifnum \x < 5
                \draw [thick, MyColor, opacity=1.0, line cap=round] (\x,\y) -- (\x+0.5,\y);
                \ifnum \y < 5
                \draw [thick, MyColor, opacity=1.0, line cap=round] (\x,\y) -- (\x+0.3,\y+0.3);
                \fi
                \ifnum \y > 1
                \draw [thick, MyColor, opacity=1.0, line cap=round] (\x,\y) -- (\x+0.3,\y-0.3);
                \fi
            \fi
            \ifnum \x > 1
                \draw [thick, MyColor, opacity=1.0, line cap=round] (\x,\y) -- (\x-0.5,\y);
                \ifnum \y < 5
                \draw [thick, MyColor, opacity=1.0, line cap=round] (\x,\y) -- (\x-0.7,\y+0.7);
                \fi
                \ifnum \y > 1
                \draw [thick, MyColor, opacity=1.0, line cap=round] (\x,\y) -- (\x-0.7,\y-0.7);
                \fi
            \fi
            \ifnum \y < 5
                \draw [thick, MyColor, opacity=1.0, line cap=round] (\x,\y) -- (\x,\y+0.5);
            \fi
            \ifnum \y > 1
                \draw [thick, MyColor, opacity=1.0, line cap=round] (\x,\y) -- (\x,\y-0.5);
            \fi
        }
    \foreach \x in {1,...,5}
        \foreach \y in {1,...,5}{
            %\filldraw[fill=black] (\x,\y) circle [radius=0.7*\r];
        }
    \foreach \x in {1,...,5}
        \foreach \y in {1,...,5}{
            \ifnum \y < 5
                \filldraw[fill=black] (\x,\y+0.5) circle [radius=\r];
                \filldraw[fill=black] (\y+0.5,\x) circle [radius=\r];
            \fi
        }
    \foreach \x in {1,...,4}
        \foreach \y in {1,...,4}{
            \filldraw[fill=black] (\x+0.3,\y+0.3) circle [radius=\r];
            \filldraw[fill=black] (\y+0.3,\x+0.7) circle [radius=\r];
        }
    \tikzset{decoration={snake,amplitude=.4mm,segment length=2mm,
                    post length=0mm,pre length=0mm}}
    \draw [decorate] (2.7, 0.5) -- (2.7, 5.5);
\end{tikzpicture}}
Vertex tensors are not shown here because they can be absorbed into an edge tensor and hence do not change the contraction complexity.
If we contract this tensor network in the column-wise order, the maximum intermediate tensor has rank $\sim L$, which can be seen by counting the number of colors at the cut.

@@

In this diagram, the additional $\delta$ tensors can have ranks up to $8$.
If we still contract this tensor network in a column-wise order, the maximum intermediate tensor has rank $\sim3L$, i.e. the space complexity is $\approx 2^{3L}$, which has a larger complexity than using the generalized tensor network notation.

\section{Hard problems and tensor networks}\label{sec:otherproblems}
\subsection{Maximal independent sets and maximal cliques}\label{sec:maximal}
In this section, we focus the discussion on the maximal independent sets problem since finding maximal cliques of a graph is equivalent to finding the maximal independent sets of the complement of the graph.
Let $G=(V,E)$ be a graph, we denote the neighborhood of a vertex $v\in V$ as $N(v)$.
A maximal independent set $I_m$ is an independent set such that no $v \in V$ satisfies $I_m \cap (\{v\} \cup N[v])  = \emptyset$, i.e.\ an independent set that cannot become a larger one by adding a new vertex.
To characterize the maximal independence restriction, we defined a tensor on each $\{v\}\cup N(v)$ as
\begin{equation}\label{eq:maximal}
    T(x_v, w_v)_{s_1,s_2,\ldots,s_{|N(v)|},s_v} = \begin{cases}
        s_vx_v^{w_v} & s_1=s_2=\ldots=s_{|N(v)|}=0,\\
        1-s_v& \text{otherwise}.\\
    \end{cases}
\end{equation}
It means if none of the neighbours of $v$ are in $I_{m}$ ($ s_1=s_2=\ldots=s_{|N(v)|}=0$), then $v$ must be in $I_{m}$ and contribute a factor $x_{v}$, otherwise, if any of the neighbourhood vertices is in $I_{m}$, then $v$ cannot be in $I_{m}$.
For a vertex $v$ that has a degree 2, this tensor has the following form
\begin{equation}
    T(x_v, w_v)=\left(\begin{matrix}
    \left(\begin{matrix}
        ~~0 &~1 \\
        ~~1 &~1
    \end{matrix}\right)\\
    \left(\begin{matrix}
        x_v^{w_v} &0 \\
        0 &0
    \end{matrix}\right)
    \end{matrix}\right).
\end{equation}

Let us consider the graph in Example~\ref{eg:tensorcontraction}. The corresponding tensor network structure for computing the maximal independent polynomial has the following hypergraph representation. \\

@@

%The contraction complexity of this tensor network is often greater than the one for the independence set problem.
By contracting this tensor network with generic element types,
we can compute the maximal independent set properties such as the maximal independence polynomial and the enumeration of maximal independent sets.
The maximal independence polynomial is defined as
\begin{equation}
D_{m}(G, x) = \sum_{k=0}^{\alpha(G)} b_k x^k,
\end{equation}
where $b_k$ is the number of maximal independent sets of size $k$.
Comparing with the independence polynomial in \Eq{eq:idpdef}, we have $b_{k} \leq a_{k}$ and $b_{\alpha(G)} = a_{\alpha(G)}$. $D_m(G, 1)$ counts the total number of maximal independent sets~\cite{Gaspers2012, Manne2013};
to our knowledge, the best algorithm has a time complexity $O(1.3642^{|V|})$~\cite{Gaspers2012}.
%If we want to find an MIS, $b_{k}$ can, in some cases, provide hints on the difficulty of finding the MIS using local algorithms~\cite{Ebadi2022}.
%The uni-modality, log-concavity, and real-rootness properties of the maximal independence polynomial for special classes of graphs have also been studied~\cite{Hu2017}.

We show the benchmark of computing the maximal independent set properties on $3$-regular graphs in \Fig{fig:benchmark-maximal},
including a comparison to the Bron-Kerbosch algorithm from Julia package \href{https://github.com/JuliaGraphs/Graphs.jl}{Graphs}~\cite{Graphs}.
\Fig{fig:benchmark-maximal}(a) shows the space and time complexities of tensor contraction, which are typically larger than those for the independent set problem.
In \Fig{fig:benchmark-maximal}(b), one can see counting maximal independent sets are much more efficient than enumerating them, while our generic tensor network approach runs slightly faster than the Bron-Kerbosch approach in enumerating all maximal independent sets.

\begin{figure} 
    \centering
    \includegraphics[width=\textwidth, trim={0cm 0cm 0cm 0cm}, clip]{figures/fig2.pdf}
    \caption{Benchmarks of computing different solution space properties of the maximal independent sets (ISs) problem on random three regular graphs at different sizes.
    (a) time and space complexity of tensor network contraction. 
    (b) The wall clock time for counting and enumeration of maximal ISs. % The missing data is due to the memory limit.
    }
    \label{fig:benchmark-maximal}
\end{figure}

@@

\subsection{Vertex coloring}
Let $G=(V,E)$ be a graph. A vertex coloring is an assignment of colors to each vertex $v\in V$ such that no edge connects two identically colored vertices. 
In a $k$-coloring problem, the number of colors is limited to less or equal to $k$.
Let us use the 3-coloring problem as an example to show how to reduce it to tensor contractions.
We first map a vertex $v \in V$ to a degree of freedom $c_v\in\{0,1,2\}$.
Then we define a tensor labeled by $c_v$ for each vertex $v$ as
\begin{equation}
    W = \left(\begin{matrix}
        r_v\\
        g_v\\
        b_v
    \end{matrix}\right),
\end{equation}
and a bond tensor labelled by $(c_u, c_v) $ for each edge $(u, v) \in E$ as
\begin{equation}
    B(x^{w_{uv}}) = \left(\begin{matrix}
        0 & x^{w_{uv}} & x^{w_{uv}}\\
        x^{w_{uv}} & 0 & x^{w_{uv}}\\
        x^{w_{uv}} & x^{w_{uv}} & 0
    \end{matrix}\right),
\end{equation}
where $r_v$, $g_v$ and $b_v$ are colors for labeling the configurations. $B$ tensors are for specifying the coloring constraints and $W$ tensors are for labeling the solutions.
%The number of possible colorings can be obtained by contracting this tensor network by setting $r_v, g_v$ and $b_v$ to $1$.
Let $x^{w_{uv}} = x$ and $r_v = g_v = b_v = 1$; we then have a graph polynomial,
in which the $k$-th coefficient is the number of coloring that $k$ bonds satisfy constraints.
If a graph is colorable, the maximum order of this polynomial should be equal to the number of edges in this graph.
Similarly, one can define an edge coloring problem by defining the tensor network on the line graph of $G$.

\subsection{Cutting problem}
In graph theory, a cut is a partition of the vertices of a graph into two disjoint subsets,
which is also known as the spin-glass problem in statistical physics.
Let $G=(V,E)$ be a graph. We associate a weight $w_v$ to each $v\in V$. To reduce the cutting problem on $G$ to the contraction of a tensor network, we first define a Boolean degree of freedom $s_v\in\{0, 1\}$ for each vertex $v\in V$.
Then for each edge $(u,v)\in E$, we define an edge matrix labeled by $(s_u, s_v)$ as
\begin{equation}
    B(x^{w_{uv}}_u, x^{w_{uv}}_v) = \left(\begin{matrix}
        1 & x_{v}^{w_{uv}}\\
        x_{u}^{w_{uv}} & 1
    \end{matrix}\right),
\end{equation}
where variables $x_u^{w_{uv}}$ and $x_v^{w_{uv}}$ are for a cut on this edge or a domain wall in a spin glass problem.
Let $x_u^{w_{uv}} = x_v^{w_{uv}} = x$; we have a graph polynomial similar to the previous ones,
in which the $k$th coefficient is two times the number of cut configurations that have size $k$ (i.e.\ cutting $k$ edges).

\subsection{Dominating Set}
In graph theory, a dominating set for a graph $G = (V, E)$ is a subset $D \subseteq V$ such that every vertex not in $D$ is adjacent to at least one member of $D$.
To reduce this problem to the contraction of a tensor network, we first map a vertex $v\in V$ to a Boolean degree of freedom $s_v\in\{0, 1\}$.
Then for each vertex $v$, we define a tensor on its closed neighborhood $\{v\} \cup N(v)$ as
\begin{equation}
T(x^{w_v}_v)_{s_1,s_2,\ldots,s_{|N(v)|},s_v} = \begin{cases}
    0 & s_1=s_2=\ldots=s_{|N(v)|}=s_v=0,\\
    1 & s_v=0,\\
    x^{w_v}_v & \text{otherwise},
\end{cases}
\end{equation}
where $w_v$ is the weight associated with the vertex $v$.
This tensor implies a configuration having a closed neighborhood of $v$ not in $D$ ($s_1=s_2=\ldots=s_{|N(v)|}=s_v=0$) cannot be a dominating set.
Otherwise, if $v$ is in $D$, this tensor contributes a multiplicative factor $x_v^{w_v}$ to the output.
The graph polynomial for the dominating set problem is known as the domination polynomial~\cite{Alikhani2009}
\begin{equation}
D(G, x) = \sum_{k=0}^{\gamma(G)} d_k x^k,
\end{equation}
where $d_k$ is the number of dominating sets of size $k$.
